Supplemental Material for: The evolution of gene expression in the term placenta of viviparous mammals

### 2016-01-28 manuscript/manuscript.Rnw
move RLN2, OVOS2, LGAL1, LGALS13 trees to supplmenetal

To determine enrichment levels of functional categories of genes in
highly-expressed transcripts in each transcriptome, we performed
functional annotation cluster analyses per transcriptome using topGO
\cite{topGO:2.20.0}. Gene Ontology clusters were considered
significantly enriched if the \ac{FDR} as estimated by the Benjamini
and Hochberg procedure \cite{Benjamini1995:bh_procedure} on from the
$p$ values obtained from the Fisher Exact Test were less than or equal
to 0.05. Enrichment tests were considered with two gene lists per
transcriptome: annotated genes with expression levels of the top
5\% of the frequency distribution of \ac{FPKM} (Supplemental
\cref{tab:go_terms_enriched_top_five_dasypus_novemcinctus,tab:go_terms_enriched_top_five_equus_caballus,tab:go_terms_enriched_top_five_homo_sapiens,tab:go_terms_enriched_top_five_loxodonta_africana,tab:go_terms_enriched_top_five_monodelphis_domestica,tab:go_terms_enriched_top_five_mus_musculus,tab:go_terms_enriched_top_five_ovis_aries,tab:go_terms_enriched_top_five_pan_paniscus,tab:go_terms_enriched_top_five_sus_scrofa,tab:go_terms_enriched_top_five_ateles_fusciceps,tab:go_terms_enriched_top_five_nannospalax_galili,tab:go_terms_enriched_top_five_spalax_carmeli,tab:go_terms_enriched_top_five_bos_taurus,tab:go_terms_enriched_top_five_canis_familiaris});
the top 100 genes with the highest \ac{FPKM} (Supplemental
\cref{tab:go_terms_enriched_dasypus_novemcinctus,tab:go_terms_enriched_equus_caballus,tab:go_terms_enriched_homo_sapiens,tab:go_terms_enriched_loxodonta_africana,tab:go_terms_enriched_monodelphis_domestica,tab:go_terms_enriched_mus_musculus,tab:go_terms_enriched_ovis_aries,tab:go_terms_enriched_pan_paniscus,tab:go_terms_enriched_sus_scrofa,tab:go_terms_enriched_ateles_fusciceps,tab:go_terms_enriched_nannospalax_galili,tab:go_terms_enriched_spalax_carmeli,tab:go_terms_enriched_bos_taurus,tab:go_terms_enriched_canis_familiaris}).
Due to the paucity of annotated gene ontology terms for some species,
gene lists were compared using the \human Gene Ontology
using the 1:1 \human ortholog obtained

@@

\begin{figure}
<<rln2.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="RLN2" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
@ 
\caption{\genetreeexplanation{\textit{RLN2}}}
\label{fig:rln2_evolution_tree}
\end{figure}

@@

The top 50 expressed genes in the term placenta of \armadillo are
shown in \cref{tab:top50_armadillo_placenta} with a list of all genes
with $\mathrm{FPKM}\ge 50$ in Supplemental
\cref{tab:all_dasypus_novemcinctus_placenta}. The gene with the
highest expression is
\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000045029",complete=TRUE)}
which is most similar to the \mouse \textit{Pga5} is expanded in
\opossum into three members which are expressed in placenta (the other
two being
\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000033885",complete=TRUE)}
and
\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000040431",complete=TRUE)}),
though the expansion is less than that seen \cow and \sheep
(\cref{fig:pag2_evolution_tree}). There is also high expression of
hemoglobin subunit beta,
\Sexpr{gene.expression("dasypus_novemcinctus","HBB",complete=TRUE)},
an un-annotated gene with high similarity to \human and \cow
\textit{OVOS2}
\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000033320",complete=TRUE)}
(\cref{fig:ovos2_evolution_tree}), the growth factors
\Sexpr{gene.expression("dasypus_novemcinctus","IGF2",complete=TRUE)}
and
\Sexpr{gene.expression("dasypus_novemcinctus","IGFBP4",complete=TRUE)},
and the thymosin
\Sexpr{gene.expression("dasypus_novemcinctus","TMSB4X",complete=TRUE)}.
As has been seen previously, multiple components of the annexin
complex are also highly expressed, including
\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000045634",complete=TRUE)}
(\textit{ANXA8L1}),
\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000017399",complete=TRUE)}
(\textit{S100B}),
\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000048504",complete=TRUE)}
(\textit{S100A10}), and
\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000004237",complete=TRUE)}
(\textit{ANXA2}).

\begin{figure}
<<ovos2.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=

tree.id <-
    genes.to.tree["ENSDNOG00000033320",tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
                 subtree=c("ENSDNOG00000033320","ENSMODG00000017988"))
@ 
\caption{\genetreeexplanation{\cow \textit{OVOS2}}}
\label{fig:ovos2_evolution_tree}
\end{figure}

@@

\begin{table}
  \footnotesize
<<top.50.opossum.expressed.genes,echo=FALSE,results="asis">>=
top.opossum.genes <- top.genes.table(species.name="monodelphis domestica")
print(xtable(top.opossum.genes),
      booktabs=TRUE,
      floating=FALSE,
      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
setkey(top.opossum.genes,"Gene symbol")
@ 
\caption{\tableexplanation{\opossum}{tab:all_monodelphis_domestica_placenta}}
\label{tab:top50_opossum_placenta}
\end{table}

@@

In order to identify major functional categories of genes expressed in
the individual transcriptomes, we examined \ac{GSE} in 1:1 human
orthologs which had \ac{FPKM} values in the top 5\% of expressed genes
(Supplemental %
\cref{tab:go_terms_enriched_top_five_dasypus_novemcinctus,tab:go_terms_enriched_top_five_equus_caballus,tab:go_terms_enriched_top_five_homo_sapiens,tab:go_terms_enriched_top_five_loxodonta_africana,tab:go_terms_enriched_top_five_monodelphis_domestica,tab:go_terms_enriched_top_five_mus_musculus,tab:go_terms_enriched_top_five_ovis_aries,tab:go_terms_enriched_top_five_pan_paniscus,tab:go_terms_enriched_top_five_sus_scrofa,tab:go_terms_enriched_top_five_ateles_fusciceps,tab:go_terms_enriched_top_five_nannospalax_galili,tab:go_terms_enriched_top_five_spalax_carmeli,tab:go_terms_enriched_top_five_bos_taurus,tab:go_terms_enriched_top_five_canis_familiaris}) using topGO \cite{topGO:2.20.0}. %
%
Significant enrichment in genes ($\mathrm{FDR}\le0.05$) in the top 5\%
of \ac{FPKM} values in at least 12 of the
\Sexpr{length(species.names)} species included clusters such as
membrane organization , locomotion, tissue development, extracellular
matrix organization, regulation of peptidase activity, and various
cell death related pathways among others
(\cref{tab:go_terms_all_species}). If we instead examine the top 100
genes by \ac{FPKM} in each of the \Sexpr{length(species.names)}
species (Supplemental %
\cref{tab:go_terms_enriched_dasypus_novemcinctus,tab:go_terms_enriched_equus_caballus,tab:go_terms_enriched_homo_sapiens,tab:go_terms_enriched_loxodonta_africana,tab:go_terms_enriched_monodelphis_domestica,tab:go_terms_enriched_mus_musculus,tab:go_terms_enriched_ovis_aries,tab:go_terms_enriched_pan_paniscus,tab:go_terms_enriched_sus_scrofa,tab:go_terms_enriched_ateles_fusciceps,tab:go_terms_enriched_nannospalax_galili,tab:go_terms_enriched_spalax_carmeli,tab:go_terms_enriched_bos_taurus,tab:go_terms_enriched_canis_familiaris})
notable enrichments include extracellular matrix, female pregnancy,
integrin and collagen-related processes, and various cell death
related processes.

@@

Relaxin (\textit{RLN}) is a hormone that has multiple effects during
pregnancy including increasing maternal blood flow, widening the pubic
bone and relaxing uterine musculature in preparation for labor. In
\dog, \elephant, \mouse, \carmeli, and \horse there is high expression
of relaxin in the term placenta, with little evidence in other species
examined here (\cref{fig:rln2_evolution_tree}). \textit{RLN} is known
to be produced in the placenta of the pregnant \dog and is used in the
diagnosis of canine pregnancy
\cite{Steinetz.ea1989:Serumrelaxinandprogesterone,Tsutsui.ea1991:Determinationofsourceof},
the luteal ovary and placenta of \pig
\cite{Hisaw.ea1948:Relaxininovaryof}, and the placenta of \horse
\cite{Klonisch.ea1997:Placentallocalizationofrelaxin}. However, the
source of relaxin produced by pregnant \elephant remains unclear
\cite{Meyer.ea2004:Speciesandfetalgender}. The high rise in relaxin in
\elephant has been suggested to be a major signal for the initiation
of labor \cite{Niemuller.ea1998:Plasmaconcentrationsofimmunoreactive}.
Here we show that the placenta of elephants show high expression of
relaxin and is likely a major source of \textit{RLN} at term. In
contrast to the results of
\cite{Evans.ea1993:mouserelaxingenenucleotide}, we also see expression
of \textit{RLN} in the placenta of \mouse and \carmeli.

@@

\begin{figure}
<<nppb.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="NPPB" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
@ 
\caption{\genetreeexplanation{\textit{NPPB}}}
\label{fig:nppb_evolution_tree}
\end{figure}

\begin{figure}
<<nppc.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="NPPC" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
@ 
\caption{\genetreeexplanation{\textit{NPPC}}}
\label{fig:nppc_evolution_tree}
\end{figure}

Two genes, \textit{NPPB} and \textit{NPPC}, code for peptide hormones
that are important in the inducement of vasodilation, and may be
important in increasing maternal blood flow to the fetus
\cite{Cameron.ea1996:sitesofgeneexpression}. There is some indication
that these proteins are present in the decidua of mice
\cite{Cameron.ea1996:sitesofgeneexpression}, but we find very little
expression in the fetal portion of the mouse placenta
(\cref{fig:hormone_genes,fig:nppc_evolution_tree,fig:nppb_evolution_tree}).
However, we do find high expression of at least one of these genes in
\dog, \cow, and \elephant. The lack of expression in the hemochorial
placentas of Euarchontoglires may mean that high expression of these
genes are essential in increasing maternal blood flow in
epitheliochorial and endotheliochorial placentas.

@@

\begin{figure}
<<galectins.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="LGALS1" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
@ 
\caption{Expression of galectins}
\label{fig:galectins_evolution_tree}
\end{figure}

\begin{figure}
<<lgals13.galectins.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="LGALS13" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
@ 
\caption{\genetreeexplanation{\textit{LGALS13}}}
\label{fig:galectins_lgals13_evolution_tree}
\end{figure}
%$
Galectins are recognized as major regulators of pregnancy in both
fetal and maternal tissues
\cite{Than.ea2012:Galectinsguardiansofeutherian}. Indeed, the
evolution of new lineage-specific galectins has occurred in both
primates and ruminants, with the genes \textit{LGALS13},
\textit{LGALS14}, and \textit{LGALS16} recognized as being
placenta-specific in \human and regulating immune responses by
inducing apoptosis of maternal T-cells
\cite{Than.ea2009:primatesubfamilyofgalectins}. Indeed,
\textit{LGALS14} has the highest expression of any galectin in the
\human and \smonkey placenta
(\cref{fig:hormone_genes,fig:galectins_lgals13_evolution_tree}), and
other primate galectins such as \textit{LGALS13} and \textit{LGALS16}
are also expressed in \human and \smonkey placenta. However, in the
closely related \bonobo, \textit{LGALS14} is not found, and
\textit{LGALS1} is the most abundant galectin. \textit{LGALS1} shows
generally high expression in all placentas analyzed. \textit{LGALS1}
is also involved in mediating maternal-fetal immune tolerance, and is
expressed in both maternal and fetal tissue in primates
\cite{Than.ea2008:Emergenceofhormonaland}. \textit{LGALS1} placental
expression was thought to be confined to eutherian mammals, as there
is evidence of adaptive evolution at this node, as well as the
appearance of estrogen response elements in eutherian cis-regulatory
regions \cite{Than.ea2008:Emergenceofhormonaland}. It was unclear as
to whether galectins such as \textit{LGALS1} were important in the
regulation of pregnancy in marsupials, as well as eutherians. Here we
show substantial expression of galectins in the opossum, especially
\textit{LGALS1} (Table 8). In addition, \textit{LGALS2} is also
expressed. The presence of galectins at the maternal-fetal interface
in \opossum implies that it was also present in the ancestral placenta
of therian mammals. Indeed, adaptive evolution of \textit{LGALS1} on
the eutherian stem is likely due to the increased invasive nature of
eutehrian placentation rather than the origination of maternal-fetal
tolerance. Although the function of \textit{LGALS1} in the marsupial
placenta needs to be confirmed, it is likely that the mediation of
maternal-fetal immune tolerance by galectins is an ancient feature of
mammalian viviparity.

>>>>

To determine enrichment levels of functional categories of genes in
highly-expressed transcripts in each transcriptome, we performed
functional annotation cluster analyses per transcriptome using topGO
\cite{topGO:2.20.0}. Gene Ontology clusters were considered
significantly enriched if the \ac{FDR} as estimated by the Benjamini
and Hochberg procedure \cite{Benjamini1995:bh_procedure} on from the
$p$ values obtained from the Fisher Exact Test were less than or equal
to 0.05. Enrichment tests were considered with two gene lists per
transcriptome: annotated genes with expression levels of the top
5\% of the frequency distribution of \ac{FPKM} (Supplemental
\cref{tab:go_terms_enriched_top_one_dasypus_novemcinctus,tab:go_terms_enriched_top_one_equus_caballus,tab:go_terms_enriched_top_one_homo_sapiens,tab:go_terms_enriched_top_one_loxodonta_africana,tab:go_terms_enriched_top_one_monodelphis_domestica,tab:go_terms_enriched_top_one_mus_musculus,tab:go_terms_enriched_top_one_ovis_aries,tab:go_terms_enriched_top_one_pan_paniscus,tab:go_terms_enriched_top_one_sus_scrofa,tab:go_terms_enriched_top_one_ateles_fusciceps,tab:go_terms_enriched_top_one_nannospalax_galili,tab:go_terms_enriched_top_one_spalax_carmeli,tab:go_terms_enriched_top_one_bos_taurus,tab:go_terms_enriched_top_one_canis_familiaris});
the top 100 genes with the highest \ac{FPKM} (Supplemental
\cref{tab:go_terms_enriched_dasypus_novemcinctus,tab:go_terms_enriched_equus_caballus,tab:go_terms_enriched_homo_sapiens,tab:go_terms_enriched_loxodonta_africana,tab:go_terms_enriched_monodelphis_domestica,tab:go_terms_enriched_mus_musculus,tab:go_terms_enriched_ovis_aries,tab:go_terms_enriched_pan_paniscus,tab:go_terms_enriched_sus_scrofa,tab:go_terms_enriched_ateles_fusciceps,tab:go_terms_enriched_nannospalax_galili,tab:go_terms_enriched_spalax_carmeli,tab:go_terms_enriched_bos_taurus,tab:go_terms_enriched_canis_familiaris}).
Due to the paucity of annotated gene ontology terms for some species,
gene lists were compared using the \human Gene Ontology
using the 1:1 \human ortholog obtained

@@

@@

The top 50 expressed genes in the term placenta of \armadillo are
shown in \cref{tab:top50_armadillo_placenta} with a list of all genes
with $\mathrm{FPKM}\ge 50$ in Supplemental
\cref{tab:all_dasypus_novemcinctus_placenta}. The gene with the
highest expression is
\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000045029",complete=TRUE)}
which is most similar to the \mouse \textit{Pga5} is expanded in
\opossum into three members which are expressed in placenta (the other
two being
\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000033885",complete=TRUE)}
and
\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000040431",complete=TRUE)}),
though the expansion is less than that seen \cow and \sheep
(\cref{fig:pag2_evolution_tree}). There is also high expression of
hemoglobin subunit beta,
\Sexpr{gene.expression("dasypus_novemcinctus","HBB",complete=TRUE)},
an un-annotated gene with high similarity to \human and \cow
\textit{OVOS2}
\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000033320",complete=TRUE)}
(Supplemental \cref{fig:ovos2_evolution_tree}), the growth factors
\Sexpr{gene.expression("dasypus_novemcinctus","IGF2",complete=TRUE)}
and
\Sexpr{gene.expression("dasypus_novemcinctus","IGFBP4",complete=TRUE)},
and the thymosin
\Sexpr{gene.expression("dasypus_novemcinctus","TMSB4X",complete=TRUE)}.
As has been seen previously, multiple components of the annexin
complex are also highly expressed, including
\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000045634",complete=TRUE)}
(\textit{ANXA8L1}),
\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000017399",complete=TRUE)}
(\textit{S100B}),
\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000048504",complete=TRUE)}
(\textit{S100A10}), and
\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000004237",complete=TRUE)}
(\textit{ANXA2}).

@@

\begin{table}
  \footnotesize
<<top.50.opossum.expressed.genes,echo=FALSE,results="asis">>=
output.top.genes.table(top.genes.table(species.name="monodelphis domestica"))
@ 
\caption{\tableexplanation{\opossum}{tab:all_monodelphis_domestica_placenta}}
\label{tab:top50_opossum_placenta}
\end{table}

@@

In order to identify major functional categories of genes expressed in
the individual transcriptomes, we examined \ac{GSE} in 1:1 human
orthologs which had \ac{FPKM} values in the top 1\% of expressed genes
(Supplemental %
\cref{tab:go_terms_enriched_top_one_dasypus_novemcinctus,tab:go_terms_enriched_top_one_equus_caballus,tab:go_terms_enriched_top_one_homo_sapiens,tab:go_terms_enriched_top_one_loxodonta_africana,tab:go_terms_enriched_top_one_monodelphis_domestica,tab:go_terms_enriched_top_one_mus_musculus,tab:go_terms_enriched_top_one_ovis_aries,tab:go_terms_enriched_top_one_pan_paniscus,tab:go_terms_enriched_top_one_sus_scrofa,tab:go_terms_enriched_top_one_ateles_fusciceps,tab:go_terms_enriched_top_one_nannospalax_galili,tab:go_terms_enriched_top_one_spalax_carmeli,tab:go_terms_enriched_top_one_bos_taurus,tab:go_terms_enriched_top_one_canis_familiaris}) using topGO \cite{topGO:2.20.0}. %
%
Significant enrichment in genes ($\mathrm{FDR}\le0.05$) in the top 5\%
of \ac{FPKM} values in at least 12 of the
\Sexpr{length(species.names)} species included clusters such as
membrane organization , locomotion, tissue development, extracellular
matrix organization, regulation of peptidase activity, and various
cell death related pathways among others
(\cref{tab:go_terms_all_species}). If we instead examine the top 100
genes by \ac{FPKM} in each of the \Sexpr{length(species.names)}
species (Supplemental %
\cref{tab:go_terms_enriched_dasypus_novemcinctus,tab:go_terms_enriched_equus_caballus,tab:go_terms_enriched_homo_sapiens,tab:go_terms_enriched_loxodonta_africana,tab:go_terms_enriched_monodelphis_domestica,tab:go_terms_enriched_mus_musculus,tab:go_terms_enriched_ovis_aries,tab:go_terms_enriched_pan_paniscus,tab:go_terms_enriched_sus_scrofa,tab:go_terms_enriched_ateles_fusciceps,tab:go_terms_enriched_nannospalax_galili,tab:go_terms_enriched_spalax_carmeli,tab:go_terms_enriched_bos_taurus,tab:go_terms_enriched_canis_familiaris})
notable enrichments include extracellular matrix, female pregnancy,
integrin and collagen-related processes, and various cell death
related processes.

@@

Relaxin (\textit{RLN}) is a hormone that has multiple effects during
pregnancy including increasing maternal blood flow, widening the pubic
bone and relaxing uterine musculature in preparation for labor. In
\dog, \elephant, \mouse, \carmeli, and \horse there is high expression
of relaxin in the term placenta, with little evidence in other species
examined here (Supplemental \cref{fig:rln2_evolution_tree}). \textit{RLN} is known
to be produced in the placenta of the pregnant \dog and is used in the
diagnosis of canine pregnancy
\cite{Steinetz.ea1989:Serumrelaxinandprogesterone,Tsutsui.ea1991:Determinationofsourceof},
the luteal ovary and placenta of \pig
\cite{Hisaw.ea1948:Relaxininovaryof}, and the placenta of \horse
\cite{Klonisch.ea1997:Placentallocalizationofrelaxin}. However, the
source of relaxin produced by pregnant \elephant remains unclear
\cite{Meyer.ea2004:Speciesandfetalgender}. The high rise in relaxin in
\elephant has been suggested to be a major signal for the initiation
of labor \cite{Niemuller.ea1998:Plasmaconcentrationsofimmunoreactive}.
Here we show that the placenta of elephants show high expression of
relaxin and is likely a major source of \textit{RLN} at term. In
contrast to the results of
\cite{Evans.ea1993:mouserelaxingenenucleotide}, we also see expression
of \textit{RLN} in the placenta of \mouse and \carmeli.

@@

Two genes, \textit{NPPB} and \textit{NPPC}, code for peptide hormones
that are important in the inducement of vasodilation, and may be
important in increasing maternal blood flow to the fetus
\cite{Cameron.ea1996:sitesofgeneexpression}. There is some indication
that these proteins are present in the decidua of mice
\cite{Cameron.ea1996:sitesofgeneexpression}, but we find very little
expression in the fetal portion of the mouse placenta
(Supplemental \cref{fig:hormone_genes,fig:nppc_evolution_tree,fig:nppb_evolution_tree}).
However, we do find high expression of at least one of these genes in
\dog, \cow, and \elephant. The lack of expression in the hemochorial
placentas of Euarchontoglires may mean that high expression of these
genes are essential in increasing maternal blood flow in
epitheliochorial and endotheliochorial placentas.

@@

Galectins are recognized as major regulators of pregnancy in both
fetal and maternal tissues
\cite{Than.ea2012:Galectinsguardiansofeutherian}. Indeed, the
evolution of new lineage-specific galectins has occurred in both
primates and ruminants, with the genes \textit{LGALS13},
\textit{LGALS14}, and \textit{LGALS16} recognized as being
placenta-specific in \human and regulating immune responses by
inducing apoptosis of maternal T-cells
\cite{Than.ea2009:primatesubfamilyofgalectins} (Supplemental
\cref{fig:galectins_lgals13_evolution_tree}). Indeed, \textit{LGALS14}
has the highest expression of any galectin in the \human and \smonkey
placenta (\cref{fig:hormone_genes}, Supplemental
\cref{fig:galectins_lgals13_evolution_tree}), and other primate
galectins such as \textit{LGALS13} and \textit{LGALS16} are also
expressed in \human and \smonkey placenta. However, in the closely
related \bonobo, \textit{LGALS14} is not found, and \textit{LGALS1} is
the most abundant galectin. \textit{LGALS1} shows generally high
expression in all placentas analyzed. \textit{LGALS1} is also involved
in mediating maternal-fetal immune tolerance, and is expressed in both
maternal and fetal tissue in primates
\cite{Than.ea2008:Emergenceofhormonaland}. \textit{LGALS1} placental
expression was thought to be confined to eutherian mammals, as there
is evidence of adaptive evolution at this node, as well as the
appearance of estrogen response elements in eutherian cis-regulatory
regions \cite{Than.ea2008:Emergenceofhormonaland}. It was unclear as
to whether galectins such as \textit{LGALS1} were important in the
regulation of pregnancy in marsupials, as well as eutherians. Here we
show substantial expression of galectins in the opossum, especially
\textit{LGALS1} (Supplemental \cref{fig:galectins_evolution_tree}). In
addition, \textit{LGALS2} is also expressed. The presence of galectins
at the maternal-fetal interface in \opossum implies that it was also
present in the ancestral placenta of therian mammals. Indeed, adaptive
evolution of \textit{LGALS1} on the eutherian stem is likely due to
the increased invasive nature of eutehrian placentation rather than
the origination of maternal-fetal tolerance. Although the function of
\textit{LGALS1} in the marsupial placenta needs to be confirmed, it is
likely that the mediation of maternal-fetal immune tolerance by
galectins is an ancient feature of mammalian viviparity.

### 2016-01-28 manuscript/supplemental_material.Rnw
move RLN2, OVOS2, LGAL1, LGALS13 trees to supplmenetal

\subsection{Top 5\% Transcripts Enrichment}

@@

@@

>>>>

\subsection{Top 1\% Transcripts Enrichment}

@@

\begin{figure}
<<rln2.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="RLN2" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
@ 
\caption{\genetreeexplanation{\textit{RLN2}}}
\label{fig:rln2_evolution_tree}
\end{figure}

\begin{figure}
<<ovos2.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=

tree.id <-
    genes.to.tree["ENSDNOG00000033320",tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
                 subtree=c("ENSDNOG00000033320","ENSMODG00000017988"))
@ 
\caption{\genetreeexplanation{\cow \textit{OVOS2}}}
\label{fig:ovos2_evolution_tree}
\end{figure}

@@

\begin{figure}
<<nppb.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="NPPB" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
@ 
\caption{\genetreeexplanation{\textit{NPPB}}}
\label{fig:nppb_evolution_tree}
\end{figure}

\begin{figure}
<<nppc.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="NPPC" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
@ 
\caption{\genetreeexplanation{\textit{NPPC}}}
\label{fig:nppc_evolution_tree}
\end{figure}

\begin{figure}
<<galectins.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="LGALS1" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
@ 
\caption{\genetreeexplanation{galectins}}
\label{fig:galectins_evolution_tree}
\end{figure}

\begin{figure}
<<lgals13.galectins.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="LGALS13" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
@ 
\caption{\genetreeexplanation{\textit{LGALS13}}}
\label{fig:galectins_lgals13_evolution_tree}
\end{figure}
